---
title: "Publication plots for sampleAllocateR manuscript"
author: "John Mulvey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    number_sections: yes
    theme: flatly
    highlight: tango
  html_document:
    toc: yes
    df_print: paged
---

Much of this code is duplicated with that in the vignette, here we tweak the figures and perform simulations with a higher number of iterations to make the graphs nicer.

# packages
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(SampleAllocateR)
library(patchwork)
```


```{r}
theme_jm <- function(base_size = 6, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) %+replace%
    theme(
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(color = "black")
    )
}
```


# simualte data
```{r}
toy_data = simulate_data(n_samples = 98, block_size = 2)
```


# generate many random layouts
```{r, message=FALSE, results='hide'}
random_seeds <- sample(1:1000000, 10000, replace = FALSE)

results_list <- lapply(random_seeds, function(seed) {
  allocate_samples(toy_data,
                   batch_size = 13,
                   covariates = c("covariate1", "covariate2", "covariate3"),
                   method = "random",
                   blocking_variable = NA,
                   seed = seed)
})
```


# plot probability distribution for a single convariate
```{r, message=FALSE, results='hide'}
# plot histrogram for a single covariate: covariate1
probability_covariate1 <- lapply(seq_along(results_list), function(i) {
  data.frame(
    iteration_number = i,
    p_value = results_list[[i]][['results']] %>%
      filter(covariate == "covariate1") %>%
      pull(p_value)
  )
})
plot_data_covariate1 <- do.call(rbind, probability_covariate1)

# plot histrogram of probability that covariate1 does not differ between the batches
p_probability_dist = plot_data_covariate1 %>%
  ggplot(aes(x = p_value)) +
  geom_histogram(boundary = 0, binwidth = 0.05) +
  theme_jm() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)))

p_probability_dist
```


# plot joint probability distribution
```{r}
balance_score <- lapply(seq_along(results_list), function(i) {
  data.frame(
    iteration_number = i,
    balance_score = calculate_balance_score(results_list[[i]][['results']]$p_value, na.rm = TRUE)
  )
})

plot_data_balance_score <- do.call(rbind, balance_score)

# plot histrogram of joint probability
p_balance_score_dist = plot_data_balance_score %>%
  ggplot(aes(x = balance_score)) +
  geom_histogram(boundary = 0, binwidth = 0.05)  +
  theme_jm() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1)))

p_balance_score_dist
```


## patchwork probability distributions
```{r}
p_probability_dist + p_balance_score_dist +
  plot_annotation(tag_levels = list(c("B", "C")))

#ggsave("../results/plots/probability_distributions.pdf")
```


# compare performance
# calcualte values
```{r}
runtime_random <- system.time({
  single_layout = allocate_samples(toy_data,
                                   batch_size = 13,
                                   covariates = c("covariate1", "covariate2", "covariate3"),
                                   method = "random",
                                   blocking_variable = NA)
})

runtime_random_search <- system.time({
  best_random_layout = allocate_samples(toy_data, 
                                   batch_size = 13, 
                                   covariates = c("covariate1", "covariate2", "covariate3"),
                                   iterations = 1000,
                                   method = "best_random")
})

runtime_optimal <- system.time({
  optimal_layout = allocate_samples(toy_data, 
                                   batch_size = 13, 
                                   covariates = c("covariate1", "covariate2", "covariate3"),
                                   iterations = 300,
                                   plot_convergence = TRUE)
})
```


# plot performance
## plot joint probabilities
```{r}
plot_performance = data.frame(
  method = c("random_search", "simulated_annealing"),
  balance_score = c(calculate_balance_score(best_random_layout[['results']]$p_value),
                        calculate_balance_score(optimal_layout[['results']]$p_value))) %>%
  mutate(method = factor(method, levels = c("random_search", "simulated_annealing"))) %>%
  ggplot(aes(x = method, y = balance_score, fill = method)) +
  geom_col() +
  theme(axis.text.x = element_blank(), #element_text(angle = 45, hjust = 1), 
        axis.title.x = element_blank()) +
  theme_jm() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

plot_performance
```


## plot computational time
```{r}
plot_runtime = data.frame(method = c("random_search", "simulated_annealing"),
           time = c(sum(runtime_random_search), sum(runtime_optimal))) %>%
  mutate(method = factor(method, levels = c("random", "random_search", "simulated_annealing"))) %>%
  ggplot(aes(x = method, y = time, fill = method)) +
  geom_col() +
  #scale_fill_manual(values = c("random_search" = scales::hue_pal()(3)[2], "simulated_annealing" = scales::hue_pal()(3)[3])) +
  theme(axis.text.x = element_blank(),#element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank()) +
  theme_jm() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(y = "Runtime (s)")

plot_runtime
```


## patchwork
```{r}
plot_performance + plot_runtime +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = list(c("D", "E"))) &
  theme(legend.position = "bottom")

#ggsave("../results/plots/performance_comparison.pdf")
```


# patchwork all
```{r}
p_probability_dist = p_probability_dist  +
  theme(axis.text.x = element_text(size = 7))

p_balance_score_dist = p_balance_score_dist +
  theme(axis.text.x = element_text(size = 7))

p_probability_dist + p_balance_score_dist + plot_performance + plot_runtime +
  plot_layout(guides = "collect", ncol = 4) &
  plot_annotation(tag_levels = list(c("B", "C", "D", "E"))) &
  theme(legend.position = "bottom")

ggsave("../results/plots/single_patchwork.pdf", width = 183, height = 70, units = "mm")
```


# plot optimsation for schematic
```{r}
optimisation_plot = optimal_layout[['optimisation_data']] %>%
    ggplot(aes(x = iteration, y = objective_value)) +
      geom_line() +
      ggtitle("Optimisation data") +
      labs(x = "Iteration", y = "Balance Score") +
      theme_jm()
    
optimisation_plot

ggsave("../results/plots/optimisation_for_schematic.pdf", width = 50, height = 50, units = "mm")
```


# sessionInfo
```{r}
sessionInfo()
```











